### 2024-04-20 docs/SOSP 2024/hir-mir-opts.tex
Change section title. Move figures in results
Minor edits

\section{Overview of \Treebeard{}}

@@

\Treebeard{} takes a serialized decision tree ensemble as input (
XGBoost JSON, ONNX etc.) and automatically generates an optimized inference function
that can either target CPUs or GPUs. 
Figure \ref{Fig:CompilerStructure} shows the structure of the \Treebeard{} compiler. 
The inference computation is lowered through three intermediate representations
-- high-level IR (HIR), mid-level IR (MIR) and low-level IR (LIR). The LIR is
finally lowered to LLVM and then JIT'ed to the specified target processor. 
\Treebeard{} is built using the open-source \TreebeardOLD{} infrastructure~\cite{Treebeard}.
% Since the \TreebeardOLD{} infrastructure was originally designed to target CPUs,
% significant extensions were required to support schedule-guided compilation for CPUs and GPUs. 
The \TreebeardOLD{} infrastructure targets CPUs.
It lacks a scheduling language and does not support generating code for 
different implementation strategies. We enhance \TreebeardOLD{}  
to support schedule-guided compilation for CPUs and GPUs.
The parts \Treebeard{} that are new or significantly different compared to \TreebeardOLD{}
are shown as shaded boxes in Figure \ref{Fig:CompilerStructure}.
% However, it implements several optimizations on the HIR and MIR that 
% can be leveraged across target processors and we find that some these 
% are beneficial for GPUs as well. 
% This reuse of optimizations is possible 
% because the intermediate representations on which these optimizations are performed 
% are abstract and are designed to be target-independent. We briefly review these 
% optimizations below.

@@

Table \ref{Tab:IRSpecification} lists the operations in the three IRs. 
In HIR, the model is represented as a collection of binary trees. This abstraction
allows the implementation of optimizations that require the manipulation of the model
or its constituent trees. We extend the \TreebeardOLD{} infrastructure with loop rewrites on 
the HIR that are implemented through the scheduling language (Section \ref{sec:schedule}).
The \emph{schedule} is implemented as an MLIR attribute on the \op{predictEnsemble} operation.
We use this object to implement the automatic scheduling described in Section \ref{sec:exploring}. 
We reuse HIR transformations to reorder and pad trees from \TreebeardOLD{}.
The tree transformations that reorder and pad trees are used in conjunction 
with loop transformations like splitting to specialize inference code as the example 
in Section \ref{sec:motivation} shows. 
% Also, we enable tree padding only if the schedule 
% specifies unrolling of tree walks.
% Figure \ref{Fig:HIRExample} shows this representation for a model 
% with four trees and how the model can be transformed. The model contains two trees of 
% depth 1 and two trees of depth 2 (\circled{A}). The right side of the diagram shows the models after 
% padding and reordering (\circled{B}). Padding makes all leaves in a tree the same depth (Trees 2 and
% 4 are padded). Trees are reordered so that trees of the same depth are together (Trees 
% 2 and 3 are swapped). We use this model as a running example for the rest of the section.
% After these model-level optimizations are performed on the HIR, the 
% code is lowered to the mid-level IR (MIR) as dictated by a  
% \emph{\textbf{schedule}}. The schedule determines how to traverse
% the iteration space that goes over the trees and input rows 
% by specifying how loops are to be tiled, 
% parallelized, mapped to GPU grid and block dimensions etc.
% (Details in Section \ref{sec:schedule}). While MIR is a loop-based IR that 
% explicitly encodes details of the iteration space has to be traversed, 
% it still abstracts details about the in-memory representation of the model.

>>>>

\section{Overview of \Treebeard{} Compiler}

@@

\Treebeard{} takes a serialized decision tree ensemble as input (
XGBoost JSON, ONNX etc.) and automatically generates an optimized inference function
that can either target CPUs or GPUs. 
Figure \ref{Fig:CompilerStructure} shows the structure of the \Treebeard{} compiler. 
The inference computation is lowered through three intermediate representations
-- high-level IR (HIR), mid-level IR (MIR) and low-level IR (LIR). The LIR is
finally lowered to LLVM and then JIT'ed to the specified target processor. 
\Treebeard{} is built using the open-source \TreebeardOLD{} compiler~\cite{Treebeard}
% Since the \TreebeardOLD{} infrastructure was originally designed to target CPUs,
% significant extensions were required to support schedule-guided compilation for CPUs and GPUs. 
that targets CPUs.
\TreebeardOLD{} lacks a scheduling language and does not support code generation for 
different implementation strategies. We enhance \TreebeardOLD{}  
to support schedule-guided compilation for CPUs and GPUs.
The parts of \Treebeard{} that are new or significantly different compared to \TreebeardOLD{}
are shown as shaded boxes in Figure \ref{Fig:CompilerStructure}.
% However, it implements several optimizations on the HIR and MIR that 
% can be leveraged across target processors and we find that some these 
% are beneficial for GPUs as well. 
% This reuse of optimizations is possible 
% because the intermediate representations on which these optimizations are performed 
% are abstract and are designed to be target-independent. We briefly review these 
% optimizations below.

@@

Table \ref{Tab:IRSpecification} lists the operations in the three IRs. 
In HIR, the model is represented as a collection of binary trees. This abstraction
allows the implementation of optimizations that require the manipulation of the model
or its constituent trees. We extend \TreebeardOLD{} with loop rewrites on 
the HIR that are implemented through the scheduling language (Section \ref{sec:schedule}).
The \emph{schedule} is implemented as an MLIR attribute on the \op{predictEnsemble} operation.
We use this object to implement the automatic scheduling described in Section \ref{sec:exploring}. 
We reuse HIR transformations to reorder and pad trees from \TreebeardOLD{}.
The tree transformations that reorder and pad trees are used in conjunction 
with loop transformations like splitting to specialize inference code as the example 
in Section \ref{sec:motivation} shows. 
% Also, we enable tree padding only if the schedule 
% specifies unrolling of tree walks.
% Figure \ref{Fig:HIRExample} shows this representation for a model 
% with four trees and how the model can be transformed. The model contains two trees of 
% depth 1 and two trees of depth 2 (\circled{A}). The right side of the diagram shows the models after 
% padding and reordering (\circled{B}). Padding makes all leaves in a tree the same depth (Trees 2 and
% 4 are padded). Trees are reordered so that trees of the same depth are together (Trees 
% 2 and 3 are swapped). We use this model as a running example for the rest of the section.
% After these model-level optimizations are performed on the HIR, the 
% code is lowered to the mid-level IR (MIR) as dictated by a  
% \emph{\textbf{schedule}}. The schedule determines how to traverse
% the iteration space that goes over the trees and input rows 
% by specifying how loops are to be tiled, 
% parallelized, mapped to GPU grid and block dimensions etc.
% (Details in Section \ref{sec:schedule}). While MIR is a loop-based IR that 
% explicitly encodes details of the iteration space has to be traversed, 
% it still abstracts details about the in-memory representation of the model.
